On the link between functional traits and growth rate: meta-analysis shows effects change with plant size, as predicted

### 2016-01-20 MS.tex
Add paragraph on the extend of the knowledge gap (unbalanced observations across stages) in abstract, results and discussion

\title{On the link between functional traits and growth rate: meta-analysis shows effects change with plant size, as predicted}
\author{Ana{\"i}s Gibert\textasteriskcentered, Emma F. Gray, Mark Westoby, Ian J. Wright, Daniel S. Falster}
\affiliation{Biological Sciences, Macquarie University NSW 2109, Australia\\
\textasteriskcentered Email for correspondence: \texttt{anais.gibert@gmail.com}\\
Word count: ~5538 words (without supplementary material, Fig and ref)}
\date{}

% Todo:
% Ditch abbrvos for SLA, wood density, Hmax?
% Fix plots
%   -- major revision of code -- > lots of repetition, hard to follow
%   -- Figure 2: add * against traits used in meta-analysis
%   -- check the appendix: in particular test of heterogeneity etc. need to re-run the analysis
%   -- check the results and adequation with text of ms
% - seach on XXX, footnote
% - check refs to appendices in main text

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary 288/350}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item A plant's growth rate is seen as a central element of its ecological strategy, and as determined by its traits. Yet the literature is inconsistent about the empirical correlation between functional traits and growth, casting doubt on the capacity of some prominent traits to influence growth rate.

\item We propose that traits should influence growth in a way that depends on the size of individual plants. We outline mechanisms and hypotheses based on new theory, and test these predictions in tree species using a meta-analysis of 103 studies ($>$ 500 correlations) for five traits (specific leaf area, wood density, maximum height, seed mass and maximum assimilation rate). To capture the effects of plant size, we tested for a shift in the direction of correlation between growth rates and traits across three ontogenic stages: seedling, sapling and adult.

\item Results were consistent with predictions, despite some limitations arising from unequal number of observations across stages. Specific leaf area was correlated with relative growth rate (RGR) in seedlings but not in adult plants. Correlations of growth with wood density and assimilation rate were not affected by stage. Seed mass and maximum height were correlated with RGR only in one plant stage category: seedlings and adults, respectively. Despite a sizeable literature on traits and growth, coverage was still limited for 4 of the 5 traits. Clearly, there is scope for strengthening our understanding on the link between traits and growth through adding further empirical studies for traits and size-classes underrepresented in the literature.

@@

@@

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{King-1999,Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Size has long been known to play a central role in organismal design and function \citep{Vogel:1988ux, Vogel:2003wb}, including that of trees \citep{Farnsworth:1995im, Givnish:1995ta}. Recent theory suggests that the absence of a correlation between SLA and growth at larger sizes may in fact be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii}. As plants grow larger, a greater amount of the energy is devoted towards building and maintaining non-photosynthetic tissues \citep{Givnish:1995ta}. A model by \citet{Falster:2011ii} suggests this change shifts the benefits and costs of some trait-based trade-offs, such that the net effect on the trait on growth changes with size. While recent empirical studies support the idea that the relationship between traits and growth rates may change with size (\citealt{Iida:2014ep, Iida:2014hq}), the trend and mechanisms behind this variation remain unclear.

In this paper, we investigate for five prominent traits (specific leaf area = SLA, wood or stem tissue density = WD, maximum leaf photosynthetic capacity per unit area = Aarea, asymptotic or maximum height = Hmax, and seed mass):
\begin{enumerate}
  \item How general is the tendency for correlations between traits and growth rate to vary with plant size?
  \item Do the observed relationships conform to predictions derived from a recently developed mechanistic model \citep{Falster:2011ii} (see ``Theoretical expectations'')?
\end{enumerate}
To address these questions we used a meta-analytic approach, compiling data from diverse studies reporting correlation coefficients between traits and growth. While most studies investigating the link between traits and growth have not primarily been concerned with the effects of plant size, they have generally reported trait-growth correlations for identifiable stage classes (i.e. seedling, sapling, and adult). We therefore used plant stage as a surrogate for plant size. We focused on tree and woody species since the mechanistic model used to assess our predictions is most applicable to these growth forms. We chose to assess hypotheses via meta-analysis because there already exist a considerable number of studies with salient information, allowing us to assess our hypotheses across a much broader number of species and vegetation types than would have been possible in an experiential study. While, we anticipated a reasonably-sizeable literature on trait and growth to be available, we were uncertain about the representation of all trait-stages combinations. An additional goal of our study was therefore to identify gaps in knowledge, needed to improve our understanding of the link between trait and growth.

@@

Below we first outline equations for the new growth model, adapted \citet{Falster:2011ii}. We then use this model to generate qualitative predictions on how the effect of traits on growth rate changes with plant size. These hypotheses are summarised in Table \ref{tab:trade-offs}.

@@

\subsection*{Tests for bias in sampling}

@@

In total, 103 articles published between 1983 and 2014, and containing 551 correlations linking growth rate to any of the 19 traits across a range of plant stages were identified (Fig. \ref{fig:fig1}). Climate zones spanned from temperate regions to tropics (map in Fig. \ref{fig:figA2}). Different experiment types (laboratory, field, plantation or greenhouse), growth measurements and growth form (tree, woody or across growth form) were included. The number of species used for each correlation ranged from 2 to 300 species, with an average of 31. Given this variety, it was not surprising to find a wide range of $r$ reported for each trait. For instance, $r$ varied from -0.98 to 0.99 for SLA, and from -0.74 to 0.74 for WD (Fig. \ref{fig:figA3}). Growth form, experiment type and growth measurement type were not strong predictors in general (Fig. \ref{fig:figA6}); suggesting confounding with these factors is not a likely explanation of the major results reported.

@@

TODO: XXXX Something about limited amount of data
TODO: XXX some thing about qualitative predictions \ref{Falster-2016}

@@

Clearly there is much scope for strengthening our understanding of correlations between traits and growth through (1) adding further empirical studies for traits and size-classes with weaker representation, (2) accounting more satisfactorily for variables such as light, and (3) expressing plant sizes on continuous scales of mass or height rather than through the approximate and partly subjective categories unavoidably adopted here. Additional data are specifically required in small plants (seedling and sapling) for WD and Hmax, and in large plants (adult and sapling) for Aarea and more broadly for all traits requiring a measure of whole plant biomass (e.g. LMR, LAR, NAR). However, the existing published evidence is consistent with qualitative predictions of a simple model that assesses the effects of plant size on growth. Thus, the linkage of growth to the mechanisms outlined here can shed light on tree growth trajectories and the size-dependent effects of functional traits on the changing species growth rate rankings as individuals grow larger.

>>>>

\title{On the link between functional traits and growth rate: meta-analysis shows effects change with plant size, as predicted}
\author{Ana{\"i}s Gibert\textasteriskcentered, Emma F. Gray, Mark Westoby, Ian J. Wright, Daniel S. Falster}
\affiliation{Biological Sciences, Macquarie University NSW 2109, Australia\\
\textasteriskcentered Email for correspondence: \texttt{anais.gibert@gmail.com}\\
Word count: ~8211 words /9600 (without supplementary material, Fig and ref)}
\date{}

% Todo:
% Ditch abbrvos for SLA, wood density, Hmax?
% Fix plots
%   -- major revision of code -- > lots of repetition, hard to follow
%   -- Figure 2: add * against traits used in meta-analysis
% - seach on XXX, footnote
% - check refs to appendices in main text

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary 327/350}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item A plant's growth rate is seen as a central element of its ecological strategy, and as determined by its traits. Yet the literature is inconsistent about the empirical correlation between functional traits and growth, casting doubt on the capacity of some prominent traits to influence growth rate.

\item We propose that traits should influence growth in a way that depends on the size of individual plants. We outline mechanisms and hypotheses based on new theoritical work, and test these predictions in tree species using a meta-analysis of 103 studies ($>$ 500 correlations) for five traits (specific leaf area, wood density, maximum height, seed mass and maximum assimilation rate). We also recorded data for a list of 14 others traits commonly used in the trait literature. To capture the effects of plant size, we tested for a shift in the direction of correlation between growth rates and traits across three ontogenic stages: seedling, sapling and adult.

% Recent theoretical work ? instead of new theory?

\item Results were consistent with predictions, even with some limitations arising from unequal numbers of observation across stages. Specific leaf area was correlated with relative growth rate (RGR) in seedlings but not in adult plants. Correlations of growth with wood density and assimilation rate were not affected by stage. Seed mass and maximum height were correlated with RGR only in one plant stage category: seedlings and adults, respectively. 
\item Despite a sizeable literature on traits and growth, major knowledge gaps across size-classes still exist; for one-third of traits a majority of correlations with growth reported in the literature ($>$ 75\%) come from a single and unique size-classe. Clearly, there is scope for strengthening our understanding on the link between traits and growth through adding further empirical studies for traits and size-classes underrepresented in the literature.

% not sure to keep the end of this sentence

@@

%Clearly, there is scope for strengthening our understanding on the link between traits and growth through adding further empirical studies for traits and size-classes underrepresented in the literature.

@@

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{King-1999,Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Size has long been known to play a central role in organismal design and function \citep{Vogel:1988ux, Vogel:2003wb}, including that of trees \citep{Farnsworth:1995im, Givnish:1995ta}. As plants grow larger, a greater amount of the energy is devoted towards building and maintaining non-photosynthetic tissues \citep{Givnish:1995ta}. A recent model by \citet{Falster:2011ii} suggests this change shifts the benefits and costs of some trait-based trade-offs, such that the net effect of the trait on growth changes with size. Thus the absence of a correlation between SLA and growth at larger sizes may in fact be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii}. While recent empirical studies support the idea that the relationship between traits and growth rates may change with size (\citealt{Iida:2014ep, Iida:2014hq}), the trend and mechanisms behind this variation remain unclear.

In this paper, we investigate for five prominent traits (specific leaf area = SLA, wood or stem tissue density = WD, maximum leaf photosynthetic capacity per unit area = Aarea, asymptotic or maximum height = Hmax, and seed mass):
\begin{enumerate}
  \item How general is the tendency for correlations between traits and growth rate to vary with plant size?
  \item Do the observed relationships conform to predictions derived from a recently developed mechanistic model \citep{Falster:2011ii} (see ``Theoretical expectations'')?
\end{enumerate}
To address these questions we used a meta-analytic approach, compiling data from diverse studies reporting correlation coefficients between traits and growth. While most studies investigating the link between traits and growth have not primarily been concerned with the effects of plant size, they have generally reported trait-growth correlations for identifiable stage classes (i.e. seedling, sapling, and adult). We therefore used plant stage as a surrogate for plant size. We focused on tree and woody species since the mechanistic model used to assess our predictions is most applicable to these growth forms. 
We chose to assess our hypotheses via a meta-analysis because there already exist a considerable number of studies with salient information, allowing us to assess our hypotheses across a much broader number of species and vegetation types than would have been possible in an experiential study. While, we anticipated a reasonably-sizeable literature on trait and growth to be available, we were uncertain about the representation of all trait-stages combinations. An additional goal of our study was therefore to identify gaps in the knowledge needed to improve our understanding of the link between trait and growth.

@@

Below we first outline equations for the new growth model, adapted from \citet{Falster:2011ii}. We then use this model to generate qualitative predictions on how the effect of traits on growth rate changes with plant size. These hypotheses are summarised in Table \ref{tab:trade-offs}.

@@

In total, correlations had been reported for at least two different stages across 18 traits out of the 19 targeted (Fig. \ref{fig:fig1}). We had thought it likely that for some traits correlations would be reported only for one size class. But this limitation applied only for LAR and NARmass, where studies appeared confined to seedling stages. However, for a majority of traits the number of observations across stage was higthly unbalanced; for one-third of the 19 traits $>$ 75\% of correlations with growth reported come from a single and unique plant stage (Fig. \ref{fig:fig1}).

\subsection*{Coverage of the data and assessment of publication bias}

@@

In total, 103 articles published between 1983 and 2014, and containing 551 correlations linking growth rate to any of the 19 traits across a range of plant stages were identified. Data reported here were variable in multiple ways. Climate zones spanned from temperate regions to tropics (map in Fig. \ref{fig:figA2}). Different experiment types (laboratory, field, plantation or greenhouse), growth measurements and growth form (tree, woody or across growth form) were included. The number of species used for each correlation ranged from 2 to 300 species, with an average of 31. Given this variety, it was not surprising to find a wide range of $r$ reported for each trait. For instance, $r$ varied from -0.98 to 0.99 for SLA, and from -0.74 to 0.74 for WD (Fig. \ref{fig:figA3}). Growth form, experiment type and growth measurement type were not strong predictors in general (Fig. \ref{fig:figA6}); suggesting confounding with these factors is not a likely explanation of the major results reported.

@@

TODO: XXXX Something about limited amount of data
Despite a sizeable literature on traits and growth, there are major knowledge gaps on the link between trait and growth. The number of correlations reported for each 19 traits is generally unbalanced in favor of a particular stage, for historical or practical reasons. This means that the knowledge we have on this trait is often biased toward a particular size-classe. Additional data are specifically required in small plants (seedling and sapling) for WD and Hmax, and in large plants (adult and sapling) for Aarea and more broadly for all traits requiring a measure of whole plant biomass (e.g. LMR, LAR).

TODO: XXX some thing about qualitative predictions \ref{Falster-2016}

@@

Clearly there is much scope for strengthening our understanding of correlations between traits and growth through (1) adding further empirical studies for traits and size-classes with weaker representation, (2) accounting more satisfactorily for variables such as light, and (3) expressing plant sizes on continuous scales of mass or height rather than through the approximate and partly subjective categories unavoidably adopted here. However, the existing published evidence is consistent with qualitative predictions of a simple model that assesses the effects of plant size on growth. Thus, the linkage of growth to the mechanisms outlined here can shed light on tree growth trajectories and the size-dependent effects of functional traits on the changing species growth rate rankings as individuals grow larger.
